Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-04-25 drafts/review-responses-april-2016.tex

\begin{point}{}
``are have''
\end{point}

\reply
Fixed.

\begin{point}{\revref}
``reality'' means ``simulation'' here??
\end{point}

\reply
Well, we hoped to mean ``reality'', but perhaps ``verify that theory matches reality'' is too strong; we've adjusted this to say ``realistic, individual-based models.''.
%\alisa{I think we do mean ``reality" here, but alternate opinions welcome}
%\plr{I changed to ``realistic, individual-based models''.}

@@

\begin{point}{}
 Some appendices might be transferred to supplementary material -- the derivations are fairly standard.
\end{point}

\reply
We're not strongly against moving the appendices to supplementary material,
but we'd prefer to keep them as appendices (and thus distributed with the document)
because although the results are fairly standard, 
they might not be familiar to our readership.  
We think it is also useful to have a more detailed description of a concrete model
that falls in the domain of attraction that we discuss here.

@@

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    This manuscript presents a discussion of patterns of linkage and allele frequency clines in a hybrid zone between two populations, under a scenario of mild hybrid incompatibility. It presents deterministic equations for the distribution of ancestral haplotypes along the genome and argues that the lengths of ancestry segments can be used to infer selection.

    The idea of using the lengths of ancestry blocks to detect selection in hybrid zones is interesting, and I had not encountered it before. This article will likely be of most interest to theorists, however, as the discussion of the practical confounders of the selection test is limited. 
\end{quote}

\begin{point}{\revref}
    A serious limitation of the test is that it only detects selection against heterozygotes. They claim that this would not affect results, [previously at] l.453-456:
    \textit{``Although our assumed scenario of selection against heterozygotes at a single locus is uncommon in nature, we believe that this is unlikely to drastically influence our findings, as previous studies have demonstrated that underdominant loci share similar properties with more realistic models''}
    is unconvincing.
\end{point}

\reply
\alisa{We agree with the comments of the AE on the matter, and have added additional text to the introduction \plr{(where?)} and to the referenced section \revref.}
\plr{cite \citet{may1975gene} for this}

@@

\begin{point}{}
    From a theoretical perspective, the article makes modest but valid contributions, and  Equation 4 and 5 are potentially interesting extensions to equation (1). 
    The practical benefit of using these equations rather than forward simulations appears limited.
\end{point}

\reply
We also pondered this point, and feel that the equations contribute substantially
to building intuitive understanding for how to address haplotype lengths,
and aim to present analytical expressions and simulations so that they each support the conclusions of the other.
We believe that analytical results can allow for a deeper understanding than simulation output,
as for instance the intuitive discussion at \llname{ll:discussion_summary},
which mostly derives from the analytical model.
To aid our readers in developing this understanding, we have gone through considerable pains to map terms in our equations to the biological processes of migration, selection and recombination. 
We believe that our efforts will help build a better mechanistic understanding in our readers in addition to making them aware of our results.

@@

\begin{point}{}
    While reading the manuscript, I found it difficult to distinguish the most important observations from more tangential observations. The manuscript reads like a long list of observations, and would benefit from more focus on the main findings and more structure.
\end{point}

\reply
\yb{
    We agree that a long set of theoretical results can be tedious. We have therefore reshaped the introduction to highlight our major question ``How does selection against incompatibilities in a hybrid zone influence the length distribution of ancestry tracts?'' 
}

@@

\begin{point}{}
Since the recombination-based terms in Eqs 4 and 5 do not depend on the spatially explicit part of the model or on the selection term, I would be surprised if they had not been derived somewhere.
\end{point}

\reply
We agree that this representation has likely been derived many times. 
We turned to Hartl and Clark and found a similar derivation, which is now cited in this point in the text.

@@

\begin{point}{Eq.\ref{eqn:g_pde}}
If $a<b<\theta<0$, then I interpret $g(\theta,b) = 0$. If correct, this should be specified.  Alternately, the bounds of the integral could be modified to account for this situation.
\end{point}

\reply
Thanks for catching the mistake; we modified the bounds of the integral.

\begin{point}{}
As an example of a section that felt unstructured, consider the discussion starting at \revref.  This long paragraph starts with a long list of observations, without giving a hint as to where we are going. Beginning the paragraph with the main idea for the paragraph would be useful.
\end{point}

\reply
\plr{(comment is more general than that particular location; say something about this)}
We added an introductory sentence at \revref to provide some structure. 
We have also gone through to clarify other sections that felt unclear.

@@

\begin{point}{Figure 1}
 Plotting the haplotypes to the side of the plot would help.
\end{point}

\reply
\alisa{We have remade Figure 1, with additional information that we hope will help with interpretation. We have also related this figure to the terms in Equation 5.}

\begin{point}{\revref}
 As the theory is set up in the absence of genetic drift, and, correspondingly, the local population size is large, it is unclear how a deme with 500 individuals could become empty in a Wright-Fisher model.  The (narrow) width of the habitat also needs justification.
\end{point}

\reply
Although most of the simulations shown in the main text have high population densities,
we have run simulations with considerably lower densities, 
as shown for instance in figures \ref{sfig:condAlleleFreq_tau80_comparison}, \ref{sfig:condAlleleFreq_tau320_comparison}, and \ref{sfig:condAlleleFreq_tau1280_comparison}.
Demes are never empty for very large populations, but
we did encounter empty demes in some simulations with smaller population sizes.
\plr{add reference to more low pop density figures if added}

### 2016-04-25 drafts/hybrid_zone.tex
minor edits

\paragraph{Linked clines}
We can use this diffusion model for lineages to find expected clines in ancestry,
i.e., the expected proportion of individuals who inherit from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In the notation above,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A \given X_0 = x, \; Z_0 = z\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.) \revpoint{1}{7}
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \frac{d}{dt} q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \frac{d}{dt} q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian.
The three terms in these equations come from Brownian movement of a linages (i.e. the smoothing action of local dispersal),
the tendency of a lineage to inherit from regions where its type is more common (i.e. the net flux induced by reduced fitness in the hybrid zone),
and recombination between selected backgrounds, respectively. \alisa{I think the first and second terms may have been switched??}
\yb{We note that because recombination occurs within a deme, the recombination terms in equation \eqref{eqn:q_pde} are quite standard  \citep{HartlClark}}
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

@@

\alisa{To provide some intuition about the physical scale of this signal, we note that}
genomic regions are inherited from ancestors $T$ generations ago
in blocks of size roughly $1/T$.
There will be a long segment of $A$ ancestry
if many such adjacent blocks are all inherited from the $A$ side.
Ancestries of neighboring blocks are correlated,
due to the branching process described above. 
But, if we assume they are independent, 
then since the block surrounding location $r$ in an individual at geographic location $x$ is of ancestry $A$ with probability $q_A(x,T,r)$,
we'd expect to see, roughly, 
$q_A(x,T,r)(1+2q_A(x,T,r))/(1-q_A(x,T,r))$ 
% P(N=0) = q and if N>0 then is 1 + two independent q-geometrics with mean q/(1-q),
% so is q ( 1 + 2q/(1-q) ) = q(1+2q)/(1-q) .
consecutive blocks of ancestry $A$ about a given site unlinked to the selected site.
(This assumes, crudely, that the number of $A$-blocks on either side of the enclosing block
has a Geometric distribution with parameter $q_A(x,T,r)$.)
This implies the mean length $l(x,r)$ would be $q_A(x,T,r)(1+2q_A(x,T,r))/((1-q_A(x,T,r))T)$,
and the mean conditional length $l_A(x,r)$ would be $(1+2q_A(x,T,r))/(T(1-q_A(x,T,r))$.
Furthermore, we know that if $x=0$ or if $T$ is large and $r$ is not too small, that $q_A$ is close to $1/2$
(so $l_A(x,r) \approx 4/T$);
and for $r<1/T$ that $q_A$ looks like the selected cline.
Also, we know from the discussion above that the lineage of a \emph{selected} $A$ allele,
if it is in the region where $A$ is rare,
moves at speed roughly $\sigma \sqrt{s}$ back towards the $A$ side of the zone,
returning to the region where $A$ is common in about $x/(\sigma \sqrt{s})$ generations.
Therefore, a selected $A$ allele found on the $B$ side of the zone
should carry with it a haplotype of average length $\sigma \sqrt{s}/x$
that looks like haplotypes from the center of the zone.
This analysis suggests that $A$ haplotypes in the center of the zone should be of average length $4/T$;
this is indeed what is seen at distant sites, for instance, in Figure \ref{sfig:blocksAlongChromAncBConditioning_nonnormalized}.
Haplotypes at the selected site are expected to be longer, but still of a length proportional to $1/T$,
suggesting that the normalization in the statistic $C(x,m)$ is appropriate,
as shown in Figure \ref{Supp:adjacentBlocks},
although a numerical prediction of the value of $C(x,m)$ is elusive.
% \alisa{Above argument assumes independence between lineages. Does this make block lengths conservative? }
% PR: yes, maybe, but that correction seems like a small one, and it's not getting very quantitative predictions;
% a more serious issue is that it's ignoring that the next block has a different $r$.

@@

Using a combination of theory and simulations, we present a description of the process of cline formation and haplotype structure in a relatively young (i.e. non-equilibrium) hybrid zone.
We show that clines establish over time $1/s$, and that lineages of selected loci tend to move back towards their `ancestral home' when in a geographic region where they  are unfit. 
This occurs at speed $\sigma\sqrt{s}$. Based on this we predict, and observe in simulations, that blocks of ancestry surrounding these selected loci are longer, especially when distant from the center of the cline, than those surrounding neutral loci. 
This extends previous theoretical work on hybrid zones, which has primarily focused on stable clines in allele frequency. 
Additionally, our work suggests that the ancestry block length distribution can help detect targets of purifying selection in hybrid zones.  
The resolution of this approach is expected to scale with $1/T$, as this is the physical scale over which linked clines persist.

@@

\paragraph{The nature of selection}
Although our assumed scenario of selection against heterozygotes at a single locus is uncommon in nature, we believe that this is unlikely to drastically influence our findings, as previous studies have demonstrated that underdominant loci share similar properties with more realistic models (e.g., ecological selection \citep{Barton1989,Barton1993}).

>>>>

\paragraph{Linked clines}
We can use this diffusion model for lineages to find expected clines in ancestry,
i.e., the expected proportion of individuals who inherit from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In the notation above,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A \given X_0 = x, \; Z_0 = z\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.) \revpoint{1}{7}
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \frac{d}{dt} q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \frac{d}{dt} q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian.
The three terms in these equations come from Brownian movement of a linages (i.e. the smoothing action of local dispersal),
the tendency of a lineage to inherit from regions where its type is more common (i.e. the net flux induced by reduced fitness in the hybrid zone),
and recombination between selected backgrounds, respectively. \alisa{I think the first and second terms may have been switched??}
\yb{We note that because recombination occurs within a deme, the recombination terms in equation \eqref{eqn:q_pde} are quite standard \citep{HartlClark}.} \revpoint{1}{4}
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

@@

\alisa{To provide some intuition about the physical scale of this signal, we note that}
genomic regions are inherited from ancestors $T$ generations ago
in blocks of size roughly $1/T$. \revpoint{1}{6}
There will be a long segment of $A$ ancestry
if many such adjacent blocks are all inherited from the $A$ side.
Ancestries of neighboring blocks are correlated,
due to the branching process described above. 
But, if we assume they are independent, 
then since the block surrounding location $r$ in an individual at geographic location $x$ is of ancestry $A$ with probability $q_A(x,T,r)$,
we'd expect to see, roughly, 
$q_A(x,T,r)(1+2q_A(x,T,r))/(1-q_A(x,T,r))$ 
% P(N=0) = q and if N>0 then is 1 + two independent q-geometrics with mean q/(1-q),
% so is q ( 1 + 2q/(1-q) ) = q(1+2q)/(1-q) .
consecutive blocks of ancestry $A$ about a given site unlinked to the selected site.
(This assumes, crudely, that the number of $A$-blocks on either side of the enclosing block
has a Geometric distribution with parameter $q_A(x,T,r)$.)
This implies the mean length $l(x,r)$ would be $q_A(x,T,r)(1+2q_A(x,T,r))/((1-q_A(x,T,r))T)$,
and the mean conditional length $l_A(x,r)$ would be $(1+2q_A(x,T,r))/(T(1-q_A(x,T,r))$.
Furthermore, we know that if $x=0$ or if $T$ is large and $r$ is not too small, that $q_A$ is close to $1/2$
(so $l_A(x,r) \approx 4/T$);
and for $r<1/T$ that $q_A$ looks like the selected cline.
Also, we know from the discussion above that the lineage of a \emph{selected} $A$ allele,
if it is in the region where $A$ is rare,
moves at speed roughly $\sigma \sqrt{s}$ back towards the $A$ side of the zone,
returning to the region where $A$ is common in about $x/(\sigma \sqrt{s})$ generations.
Therefore, a selected $A$ allele found on the $B$ side of the zone
should carry with it a haplotype of average length $\sigma \sqrt{s}/x$
that looks like haplotypes from the center of the zone.
This analysis suggests that $A$ haplotypes in the center of the zone should be of average length $4/T$;
this is indeed what is seen at distant sites, for instance, in Figure \ref{sfig:blocksAlongChromAncBConditioning_nonnormalized}.
Haplotypes at the selected site are expected to be longer, but still of a length proportional to $1/T$,
suggesting that the normalization in the statistic $C(x,m)$ is appropriate,
as shown in Figure \ref{Supp:adjacentBlocks},
although a numerical prediction of the value of $C(x,m)$ is elusive.
% \alisa{Above argument assumes independence between lineages. Does this make block lengths conservative? }
% PR: yes, maybe, but that correction seems like a small one, and it's not getting very quantitative predictions;
% a more serious issue is that it's ignoring that the next block has a different $r$.

@@

Using a combination of theory and simulations, we present a description of the process of cline formation and haplotype structure in a relatively young (i.e. non-equilibrium) hybrid zone. \linelabel{ll:discussion_summary}
We show that clines establish over time $1/s$, and that lineages of selected loci tend to move back towards their `ancestral home' when in a geographic region where they  are unfit. 
This occurs at speed $\sigma\sqrt{s}$. Based on this we predict, and observe in simulations, that blocks of ancestry surrounding these selected loci are longer, especially when distant from the center of the cline, than those surrounding neutral loci. 
This extends previous theoretical work on hybrid zones, which has primarily focused on stable clines in allele frequency. 
Additionally, our work suggests that the ancestry block length distribution can help detect targets of purifying selection in hybrid zones.  
The resolution of this approach is expected to scale with $1/T$, as this is the physical scale over which linked clines persist.

@@

\paragraph{The nature of selection}
Although our assumed scenario of selection against heterozygotes at a single locus is uncommon in nature, 
we believe that this is unlikely to drastically influence our findings, 
as previous studies have demonstrated that models of underdominant selection 
share similar properties with more realistic models (e.g., ecological selection \citep{Barton1989,Barton1993}). 
\revpoint{1}{1}
\plr{(more detail here or reference intro. would like to say `indistinguishable'; who can we cite?)}
